---
title: "Traumatic injury is associated with altered adrenal and gonadal steroid synthesis in the prehospital setting"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
author: Laura Bravo
output:
  html_document: 
    code_folding: show
    theme: united
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document: 
    toc: yes
fontsize: 14pt
---


<style type="text/css">

h1.title {
  font-size: 38px;
  color: Black;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  color: DarkRed;
  text-align: center;
}
</style>

In this study, we have steroid information of 31 male patients at three different time points after trauma (<1h, 4-12h, 24-48h). Only 26 have complete steroid information for the three time points. We had also immunological information for these patients but it was taken out from the study. Clinical information such as age, MODS, LOS, ISS, Drugs, Fluids.., is also included. 34 healthy controls are also included for reference. 

Two analysis with different steroid's concentration have been performed given we had two different LOQ values given. The first measurements taken with old LOQ and the new revalidated LOQ.

**This is the one with the REVALIDATED LOQ**

***


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.path = "/Users/lxb732/Desktop/TraumaSteroids")

knitr::opts_chunk$set(fig.align='center')

#out.width='95%', fig.asp=.95

```


```{r echo=TRUE, message=FALSE, warning=FALSE}

#Loading packages
#install_github("kassambara/easyGgplot2")
require(readxl)
require(reshape2)
require(plyr)
require(lubridate)
require(stringr)
require(ggrepel)
require(data.table)
require(ggpubr)
require(readr)
require(tidyverse)
require(dplyr)
require(knitr)
require(htmltools)
require(factoextra)
require(Metrics)
require(e1071)
require(ggplot2)
require(gridExtra)
require(brglm)
require(ggstatsplot)
require(compareGroups)
require(knitr)
require(kableExtra)
require(gtsummary)
require(dunn.test)
require(broom)
require(reshape2)
require("patchwork")
require(gam)
require(mgcv)
require(ComplexHeatmap)

#this might need github too
#packageurl <- "https://cran.r-project.org/src/contrib/Archive/repr/repr_1.0.2.tar.gz"
#install.packages(packageurl, repos=NULL, type="source")

set.seed(132)
```

# Data import 

## Reference steroid's concentration (LOQ)

File with LOQ concentrations (lowest concentration at which the analyte can not only be reliably detected).It includes the parameter we used at the start, and at which the original analysis was developed (original) and the corrected/revalidated one, where many datapoints were filtered as a result.

Again this file contains all analysis performed with **REVALIDATED LOQ**

```{r}

Steroids <- read_xlsx("SteroidsLow.xlsx")
Steroids <- Steroids[complete.cases(Steroids),c(2,3,4)]
names(Steroids) <- c("variable", "LOQ_Original", "LOQ_Revalidation")

```

## Healthy Patients (controls)

34 healthy controls with steroid information. Standard ranges were also recovered and measurements outside those ranges were taken out. 

```{r, message=FALSE}

#outliers per steroid. 
#saveRDS(Names, "names.rds")

RealHealthy <- function(HealthyLong){
  
Names <- data.frame(variable = unique(filter(HealthyLong, Descriptor == "Steroid")$variable))
Names["LimitLow"] <- c("30","10","1","0.7","2","0.1","2","0.6","1","0.09","0.3","2","0","0.07","0","1", rep("0",7)) #said 1 in progesterone (second to last but then all readings would be off)
Names["LimitHigh"] <- c("100","650","2.6","3.9","13","0.7","40","4","7","0.6","40","35","10","2.4","10","20",rep("100",7)) #much less in170HP but then off too * also 11KA4 very high numbers in general - muh than trauma and reference
Names

a <- full_join(HealthyLong,Names)

a$LimitLow <- as.numeric(a$LimitLow)
a$LimitHigh <- as.numeric(a$LimitHigh)

a1 <- a %>% 
  mutate(TakeOut = value >= (LimitLow)  & value <= (LimitHigh)) #take out ten percent of readings? -- Yes

HealthyAll <- a1 %>% 
  filter( TakeOut == "TRUE") %>% 
  select(-c("LimitLow","LimitHigh", "TakeOut"))

return(HealthyAll)

}

Healthy <- read_delim("FinalHealthyPat.csv", 
    ";", escape_double = FALSE, locale = locale(decimal_mark = ",",
 grouping_mark = "."), trim_ws = TRUE) %>% 
  select(-Age) #Take out Age for healthy - should be age matched either way.

names(Healthy)<- c("ID", "Label", "Cortisone", "Cortisol", "11-KTestosterone","11-KA4", "11β-OHA4","11β-OHT", "Corticosterone", "11-deoxycortisol", "Androstenedione", "11-deoxycorticosterone", "Testosterone", "DHEA", "17-hydroxyprogesterone", "Dihydrotestosterone", "Progesterone", "DHEAS", "Cortisol/Cortisone", "DHEA/DHEAS", "Cortisol/DHEAS", "Cortisol/DHEA", "DHEA/Testosterone", "Testosterone/Cortisol", "Aldosterone") #distinguish between trauma and healthy control

HealthyLong <- reshape2::melt(Healthy, id.vars=c("ID","Label"))

HealthyLong <- HealthyLong %>% 
  mutate(Time = "HC", 
         Descriptor = "Steroid") 

HealthyLong <- RealHealthy(HealthyLong) #LOQ feeds out of this one so all fine

HealthyLongLOQ <- full_join(HealthyLong,Steroids) %>%
  replace_na(list(LOQ_Original = 0, LOQ_Revalidation = 0)) %>% 
  filter(value >= LOQ_Revalidation)

HealthyWideLOQ <- HealthyLongLOQ %>% 
  select(ID, Label, variable, value, Descriptor) %>%
  pivot_wider(names_from = variable, values_from = value)

#take out healthy patients that do not follow normal ranges (??)

```

## Non healthy Patients (trauma)

31 trauma patients with cytokine, immunological and clincal information at three different time points. 

(Lesson: pivot longer will put problems with types and classes of columns but melt wont).

```{r}

Trauma <- read_csv("FinalDatasetAprilwCytokines.csv") #non healthy patients

names(Trauma)[1] <- "ID"

TraumaLong <- reshape2::melt(Trauma, id.vars=c(1))

TakeOut <- c(grep("^[*]",TraumaLong$value),which(TraumaLong$value %in% c("ND","OOR <"))) #Take out spurious values 

TraumaLong <- TraumaLong[-c(TakeOut),] %>% 
  drop_na() %>% 
  mutate(Label = "Trauma")

TraumaLong[,c("variable","Time") ] <- str_split_fixed(TraumaLong$variable, "-", 2)

TraumaLong$Time <- gsub(" ","",TraumaLong$Time)

TraumaLong$Time <- as.factor(TraumaLong$Time)

levels(TraumaLong$Time) <- c("Clinical","T1","T2","T3")

'%ni%' <- Negate('%in%')

TraumaLong <- TraumaLong %>%
  filter(variable %ni% c("IL1ra","IL6", "IL8","IL10","GCSF", "MCP1", "TNFa","IL12","IL17", "Lymphocyte (106/L)","Monocyte (106/L)","Neutro:Lymph ratio","Neutrophil (106/L)","WBC (106/L)","IG (106/L)")) %>%
  mutate(Descriptor = ifelse(Time == "Clinical", "Clinical", "Steroid"))

TraumaLong$variable <- as.factor(TraumaLong$variable)

levels(TraumaLong$variable) <-c("11-deoxycortisol","11-KA4","11-KTestosterone","11β-OHA4","11β-OHT","17-hydroxyprogesterone", "Androstenedione","Age","Alcohol", "Aldosterone","Anaesthetic","Corticosterone","Cortisol",
"Cortisol/Cortisone","Cortisol/DHEA","Cortisol/DHEAS","Cortisone","DHEA","DHEA/DHEAS","DHEA/Testosterone","DHEAS",
                              "Dihydrotestosterone","11-deoxycorticosterone","Fluids","GCS","GoldenHour","ISS","MODS", "Lactate","LOS","Mechanisms","NISS","Outcome","Drug","Progesterone","Sex","Testosterone","Testosterone/Cortisol",
                                 "T1","T2","T3")


TraumaWide <- TraumaLong %>%
  select(- c(Label, Descriptor)) %>%
  pivot_wider(values_from = value, names_from= c(Time,variable))


TraumaLongLOQ <- full_join(TraumaLong,Steroids) %>%
  replace_na(list(LOQ_Original = 0, LOQ_Revalidation = 0)) %>% 
  filter(value >= LOQ_Revalidation)

names(TraumaLongLOQ)[1] <- "ID"

TraumaWideLOQ <- TraumaLongLOQ %>% 
  select(ID, Label, Time, variable, value) %>%
  pivot_wider(names_from = c(Time,variable), values_from = value)

```

***

# Data summary (Revalidated LOQ)


```{r}

CreateTablesTP <- function(TraumaWide,TraumaWideLOQ, TimePointOut1,TimePointOut2, TimePointIn ) {
  
  TraumaT1 <- TraumaWide %>% 
  select(-c(contains(TimePointOut1), contains(TimePointOut2), "ID")) %>% 
  #drop_na() %>%
  mutate_at(vars(starts_with(TimePointIn), "Clinical_Age", "Clinical_GCS", "Clinical_ISS", "Clinical_NISS", "Clinical_LOS", "Clinical_Lactate", "Clinical_GoldenHour" ),funs(as.numeric))

names(TraumaT1) <- str_remove_all(names(TraumaT1), TimePointIn)

TableTraumaT1 <- TraumaT1 %>%
  tbl_summary( 
              type = list(all_numeric() ~ "continuous",
                          all_logical() ~ "categorical") ,
              missing_text = "(Missing)"
  ) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") 

#

TraumaT1LOQ <- TraumaWideLOQ %>% 
  select(-c(contains(TimePointOut1), contains(TimePointOut2), "ID")) %>% 
  #drop_na() %>%
  mutate_at(vars(starts_with(TimePointIn), "Clinical_Age", "Clinical_GCS", "Clinical_ISS", "Clinical_NISS", "Clinical_LOS", "Clinical_Lactate", "Clinical_GoldenHour" ),funs(as.numeric))

names(TraumaT1LOQ) <- str_remove_all(names(TraumaT1LOQ), TimePointIn)

TableTraumaT1LOQ <- TraumaT1LOQ %>%
  tbl_summary( 
              type = list(all_numeric() ~ "continuous",
                          all_logical() ~ "categorical") ,
              missing_text = "(Missing)"
  ) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") 

return(list(TableTraumaT1,TableTraumaT1LOQ ))
  
}


TP1 <- CreateTablesTP(TraumaWide,TraumaWideLOQ, "T2_","T3_", "T1_" ) 
TP2 <- CreateTablesTP(TraumaWide,TraumaWideLOQ, "T1_","T3_", "T2_" ) 
TP3 <- CreateTablesTP(TraumaWide,TraumaWideLOQ, "T2_","T1_", "T3_" ) 
  

##########################


HealthyTable <- Healthy %>% 
  #drop_na() %>%
  select(-c(ID, Label)) %>%
  tbl_summary( 
             type = list(all_numeric() ~ "continuous",
                        all_logical() ~ "categorical") ,
              missing_text = "(Missing)"
  ) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") 

HealthyTableLOQ <- HealthyWideLOQ %>% 
  #drop_na() %>%
  select(-c(ID, Label, Descriptor)) %>%
  tbl_summary( 
              type = list(all_numeric() ~ "continuous",
                          all_logical() ~ "categorical") ,
              missing_text = "(Missing)"
  ) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**")
```

## Table 1. 

```{r}

tbl_merge(tbls = list(HealthyTableLOQ, TP1[[2]], TP2[[2]], TP3[[2]]), tab_spanner = c("**Healthy Controls**", "**Time Point <1h**", "**Time Point 4-12h**", "**Time Point 24-48h**") )

```


We have more missing values in revalidated than original, because this new revalidation has caused many measurements to become invalid (missing). To start with we had 5 missing patient values in T3 and 2 for T2. 

```{r}

AllDataLOQLong <- rbind(TraumaLongLOQ, HealthyLongLOQ)
AllDataLong <- rbind(TraumaLong, HealthyLong)
```

***

## Boxplot steroid description 

General steroid concentration changes at the three different time points 

```{r include=FALSE}

#q <- AllDataLong %>% 
#  group_by(variable,Time,Label) %>% 
#  tally() 
#


PVal2 <-  function(my_comparisons, Groups){
CN <- my_comparisons

# Dunn
pv2 <- data.frame(with(Groups, dunn.test(value,Time,kw=FALSE,method="BH",table = FALSE))[4:5]) %>% 
  separate(comparisons, c("group2", "group1")) %>%
  reshape::rename( c("P.adjusted" = "p.value" )) %>%
  filter(group2 == "HC") 

pv2a <- data.frame(with(Groups, dunn.test(value,Time,kw=FALSE,method="bonferroni", table = FALSE))[4:5]) %>% 
  separate(comparisons, c("group2", "group1")) %>%
  reshape::rename( c("P.adjusted" = "p.value" ) ) %>%
  filter(group2 == "HC")

# Wilcoxon
pv <- tidy(with(Groups, pairwise.wilcox.test(value, Time, p.adjust.method = "BH", paired = FALSE ))) %>%
  filter(group2 == "HC")  #why group2 before? 

pva <- tidy(with(Groups, pairwise.wilcox.test(value, Time, p.adjust.method = "bonferroni", paired = FALSE ))) %>%
  filter(group2 == "HC")


PVal <- function(pv_final){
  
pv_final <- pv_final[order(pv_final$group1), ] 
pv_final <- pv_final %>%
  mutate( map_signif2 = case_when(  
  p.value > 0.05 ~ "ns", 
  0.01 < p.value & p.value <= 0.05 ~ "*",
  0.001 < p.value & p.value <= 0.01 ~ "**",
  0.0001 < p.value & p.value <= 0.001 ~ "***", 
  TRUE ~ "****")) 


}

pp <- list(pv2, pv2a, pv, pva)
pl <- map(pp, PVal)

return(pl)
  }



PValPlots <- function(Groups, plFin, CN){
  
  if (Groups$variable[1] == "DHEA"){
    
    MaxVal <- 60} else{
  
  MaxVal <- as.numeric(quantile(Groups$value, probs=c(0.1, 0.25,0.5, 0.75, 0.9),na.rm=TRUE)[5])
    }
   #MaxVal <- max(Groups$value) 
   
  l <- data.frame(table(Groups$Time))
  
  bp.vals <- function(x, probs=c(0.1, 0.25,0.5,  0.75, 0.9)) {
  r <- quantile(x, probs=probs , na.rm=TRUE)
  r = c(r)
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}

   p <- ggplot(Groups, aes(x=Time, y=value, fill=Label)) +
   #geom_boxplot(position=position_dodge(0.8)) +
      stat_summary(fun.data=bp.vals, geom="boxplot", position=position_dodge(0.8)) +
 # geom_dotplot(binaxis='y', stackdir='center', position=position_dodge(0.8), stackratio=0.6, dotsize=0.6) +
   facet_wrap(. ~ variable,ncol=4,scale="free") +
      theme_ggstatsplot() +
  theme(strip.text.x = element_blank())+
 labs(title=paste0( i),size=10) + 
     scale_fill_manual(values = c("white", "grey80"))
   
   
 k <- p +
  # stat_compare_means(comparisons=my_comparisons, label = "p.signif",method = "wilcox.test",size=6) + 
   geom_signif(comparisons=CN,
              y_position = c(MaxVal + sd(Groups$value), MaxVal + 1.2*sd(Groups$value), MaxVal + 1.4*sd(Groups$value)),
               annotation= plFin$map_signif2, tip_length = 0, vjust=0.2) +
  xlab(" ") + 
   ylab(Groups$Axis) +
   scale_x_discrete(labels = c(paste0("Healthy Controls \n (n=",l[l$Var1 == "HC",2],")"), paste0("<1h \n (n=",l[l$Var1 == "T1",2],")") ,paste0("4-12h \n (n=",l[l$Var1 == "T2",2],")"),paste0("48-72h \n (n=",l[l$Var1 == "T3",2],")"))) + 
   theme(legend.position = "none",
    legend.key = element_blank(), strip.background = element_rect(colour="white", fill="black")) + 
 #geom_hline(yintercept= as.numeric(Groups$LOQ_Original[1]), color = "blue", linetype = "dashed") + 
geom_hline(yintercept= as.numeric(Groups$LOQ_Revalidation[1]), color = "coral", linetype ="dashed") + 
   geom_text(aes(x = 0.8, y= as.numeric(LOQ_Revalidation[1]) ,label = as.numeric(LOQ_Revalidation[1]), vjust = 1.5), size = 3, color = "coral") +
   scale_color_discrete(name = "Y series", labels = c("Y2", "Y1")) +  scale_y_continuous( limits = c(0, NA))
 
 }
   
```

```{r ,message=FALSE, warning=FALSE, include=FALSE}

Mineralocorticoids <- c("11-deoxycorticosterone","Corticosterone","Aldosterone")
GlucocorticoidPrec <- c("Progesterone","17-hydroxyprogesterone","11-deoxycortisol")
Glucocorticoids <- c("Cortisol", "Cortisone" )
Androgens <- c("DHEA","DHEAS","Androstenedione")
OxygenatedAndro <- c("Testosterone", "Dihydrotestosterone","11-KA4","11-KTestosterone","11β-OHA4","11β-OHT")
Ratios <- c("Cortisol/Cortisone", "Cortisol/DHEA","Cortisol/DHEAS","DHEA/DHEAS","DHEA/Testosterone","Testosterone/Cortisol")


All2 <- AllDataLong %>%  #here we try all different ones - filtering already included??
  mutate(Groups = case_when(variable %in% c("11-deoxycorticosterone","Corticosterone","Aldosterone") ~ "Mineralocorticoids", 
                            variable %in% c("Progesterone","17-hydroxyprogesterone","11-deoxycortisol") ~ "GlucocorticoidPrec", 
                            variable %in% c("Cortisol", "Cortisone" ) ~ "Glucocorticoids", 
                            variable %in% c("DHEA","DHEAS","Androstenedione") ~ "Androgens", 
                            variable %in% c("Testosterone", "Dihydrotestosterone","11-KA4","11-KTestosterone","11β-OHA4","11β-OHT") ~ "OxygenatedAndro",
                            variable %in% c("Cortisol/Cortisone", "Cortisol/DHEA","Cortisol/DHEAS","DHEA/DHEAS","DHEA/Testosterone","Testosterone/Cortisol") ~ "Ratios", TRUE ~ "No steroid"
                            ), 
         Axis = case_when(Groups == "Ratios" ~ " ", 
                          variable == "DHEAS" ~ "Concentration (μmol/L)", 
                          TRUE ~ "Concentration (nmol/L)")) %>% 
  filter(Descriptor == "Steroid")


All2 <- All2 %>% 
  full_join(.,Steroids) %>%
  replace_na(list(LOQ_Original = 0, LOQ_Revalidation = 0)) 

All2$variable <- droplevels(as.factor(All2$variable))
All2$Time <- droplevels(as.factor(All2$Time))
All2$value <- as.numeric(All2$value)
All2$Time <- fct_relevel(All2$Time, "HC", "T1", "T2", "T3")
  

my_comparisons <- list(c("HC", "T1"),c("HC", "T2"),c("HC", "T3"))
CN <- my_comparisons

k <- list()
k1 <- list()
k3 <- list()
count <- 0
All2$variable <- as.factor(as.character(All2$variable))


HypoTest <- 3

for (i in levels(All2$variable)){
  
 # for (i in "DHEA"){
  
   levels(All2$Groups) <- c("light blue", "yellow","orange","green","steelblue3","grey")
   
   count <- count + 1

   Groups <- filter(All2, variable == i)
   
   
   ########################

  pl <- (PVal2(my_comparisons,Groups))
  plFin <- pl[[HypoTest]]
  
  k[[i]] <- (PValPlots(Groups,plFin, CN))
 
 ################
 
 #LOQ
 
  
 Groups2 <- Groups %>%
   mutate(NewValOriginal = case_when(
     value < LOQ_Original ~ "low", 
     TRUE ~ "ok"
   )) %>%
   filter(NewValOriginal == "ok")
  
 pl2 <-  (PVal2(my_comparisons,Groups2))
 plFin2 <- pl2[[HypoTest]]
 
 k1[[i]] <- (PValPlots(Groups2,plFin2,CN))

 
 
 ##################
 
 Groups3 <- Groups %>%
   mutate(NewValRe = case_when(
     value < LOQ_Revalidation ~ "low", 
     TRUE ~ "ok"
   )) %>%
   filter(NewValRe == "ok")
  

pl3 <-  (PVal2(my_comparisons,Groups3))
plFin3 <- pl3[[HypoTest]]
 
 k3[[i]] <- (PValPlots(Groups3,plFin3,CN))
 
}


```


### Steroid concentration figures

```{r include=FALSE}

pdf("Fig1Trauma.pdf",  10,  13)
#plot_spacer() |plot_spacer() | plot_spacer())
(k3[["Progesterone"]]   |  k3[["17-hydroxyprogesterone"]]) / 
  (k3[["11-deoxycortisol"]]|  k3[["Cortisol"]]) / 
 (k3[["Cortisone"]] | k3[["Cortisol/Cortisone"]]) 

dev.off()

#options(repr.plot.width=18, repr.plot.height=5.5)
pdf("Fig2Trauma.pdf",  13,  5)
k3[["11-deoxycorticosterone"]] | k3[["Corticosterone"]] | k3[["Aldosterone"]] 
dev.off()

pdf("Fig3Trauma.pdf",  13,  9)


(k3[["DHEA"]] | k3[["DHEAS"]] | k3[["DHEA/DHEAS"]]) /
(k3[["Androstenedione"]] | k3[["Testosterone"]] | k3[["Dihydrotestosterone"]]) 

dev.off()

pdf("Fig4Trauma.pdf",  10,  8)

(k3[["11β-OHA4"]]  | k3[["11β-OHT"]]) /
(k3[["11-KA4"]]  | k3[["11-KTestosterone"]] )

dev.off()
```

```{r Fig1Trauma, fig.cap = "Figure.1",  fig.height= 10, fig.width= 7}

(k3[["Progesterone"]]   |  k3[["17-hydroxyprogesterone"]]) / 
  (k3[["11-deoxycortisol"]]|  k3[["Cortisol"]]) / 
 (k3[["Cortisone"]] | k3[["Cortisol/Cortisone"]]) 

```

```{r Fig2Trauma,fig.cap = "Figure.2", fig.height= 4, fig.width= 7}

k3[["11-deoxycorticosterone"]] | k3[["Corticosterone"]] | k3[["Aldosterone"]] 

```

```{r Fig3Trauma, fig.cap = "Figure.3",fig.height=8, fig.width= 9}


(k3[["DHEA"]] | k3[["DHEAS"]] | k3[["DHEA/DHEAS"]]) /
(k3[["Androstenedione"]] | k3[["Testosterone"]] | k3[["Dihydrotestosterone"]]) 

```

```{r Fig4Trauma, fig.cap = "Figure.4", fig.height=7, fig.width= 7}


(k3[["11β-OHA4"]]  | k3[["11β-OHT"]]) /
(k3[["11-KA4"]]  | k3[["11-KTestosterone"]] )

```

## Golden Hour Study

Study of steroid concentration at the first time point (<1h). Use of Generalized Additive Models (GAMs) to extract trajectories through time. 

```{r include=FALSE}


maxVal <- 60

GAMBuild <- function(a0,type){
   
    mod.logsynd.elev.gam <- gam(log(var)~s(GHTime,k=2),data=a0,method="REML") #6
    newd<-data.frame(GHTime=seq(0,maxVal,by=0.5))
    pred.elev.synd <- predict.gam(mod.logsynd.elev.gam,newd,se.fit=T)
    pred.elev.synd$Time<-seq(0,maxVal,by=0.5)
    pred.elev.synd$Fit<-exp(pred.elev.synd$fit)
    pred.elev.synd$Lower<-exp(pred.elev.synd$fit-qnorm(0.975)*pred.elev.synd$se.fit)
    pred.elev.synd$Upper<-exp(pred.elev.synd$fit+qnorm(0.975)*pred.elev.synd$se.fit)
    pred.elev.synd$Syndecan <- type
    pred.elev.synd<-as.data.frame(pred.elev.synd)
    return(pred.elev.synd)
}

GAMModel<- function(a,i){

    print("running")
#a <- mutate(PlotGH, Unhealthy = ifelse(var >= as.numeric((quantile(filter(PlotGH, Label == "Healthy")$var,probs=c(.025,.97))[2])),"yes","no"))
a2 <- filter(a, GHTime == 0)
a2["Unhealthy"] <- "yes"
a <- rbind(a,a2)

a$Unhealthy <- as.factor(a$Unhealthy)

a0 <- filter(a, Unhealthy == "yes")
a0 %>% arrange(Label) 

pred.elev.synd <- GAMBuild(a0,"Elevated Range")

a1 <- filter(a, Unhealthy == "no")
a1 %>% arrange(Label) 

pred.norm.synd <- GAMBuild(a1,"Normal Range")

pred.synd <-rbind(pred.elev.synd,pred.norm.synd)


synd.max.plus2<-quantile(subset(a,GHTime==0,var)[,1],c(0.975))
synd.min.plus2<-quantile(subset(a,GHTime==0,var)[,1],c(0.025))

PlotGH <- PlotGH %>% mutate(Third = 
  case_when(Label == "Healthy" ~ "Healthy controls", 
            Unhealthy == "no" ~ "Non-elevated trauma patients", 
            TRUE ~ "Elevated trauma patients")
)



synd.plot2<-ggplot(data=pred.synd,aes(x=Time,y=Fit,group=Syndecan,fill=Syndecan)) + theme_bw(base_size=8) + ggtitle(paste0(i)) + xlab('Time (minutes)') + ylab(i)
synd.plot2<-synd.plot2 + geom_ribbon(aes(ymax=synd.max.plus2,ymin=synd.min.plus2,fill="Healthy Limits"),col=NA,alpha=0.05) 
synd.plot2<-synd.plot2 + geom_ribbon(aes(ymin=Lower,ymax=Upper),alpha=0.25) + geom_line(size=1) + scale_x_continuous(limits = c(0,maxVal), breaks=seq(0,maxVal,by=10), labels=seq(0,maxVal,by=10),expand = c(0.01, 0))#+ ylim(-1,60) #explore this more
synd.plot2<-synd.plot2 + geom_point(data=PlotGH,aes(x=GHTime,y=var,group=Third,colour=Third),shape=21,size=1,fill='black',stroke=0.8)

#Which days sees the mean predicted Syndecan value in elevated patients cross over the upper 2.5th percentile of the distribution of 'Normal' values?
low.lim<-pred.elev.synd$Time[which(pred.elev.synd$Lower>synd.max.plus2)[[1]]]
up.lim<-pred.elev.synd$Time[which(pred.elev.synd$Upper>synd.max.plus2)[[1]]]
est.ave<-pred.elev.synd$Time[which(pred.elev.synd$Fit>synd.max.plus2)[[1]]]
StartingTime <- c(est.ave,low.lim,up.lim)
est.line<-data.frame(x.val=c(est.ave,est.ave),y.val=c(0,synd.max.plus2))


synd.plot2 <- synd.plot2 + 
  geom_vline(aes(xintercept=est.line$x.val[1],linetype='Estimated time of onset'), col = "black",lwd=0.7) + scale_color_manual("", values = c("dark orange", "dark green", "blue"), labels = c("Trauma: Elevated levels", "Healthy controls", "Trauma: Non-elevated levels")) + scale_linetype_manual("", values = "dashed") + scale_fill_manual("", values = c("red", "green", "blue"), labels = c("Trauma: Elevated range", "Healthy baseline range", "Trauma: Non-elevated range")) + guides(fill= guide_legend(title = ""))  + theme(legend.text=element_text(size=10))


    return(list(synd.plot2,StartingTime))
}




GAMModel2<- function(a,i){

    print("running")
#a <- mutate(PlotGH, Unhealthy = ifelse(var >= as.numeric((quantile(filter(PlotGH, Label == "Healthy")$var,probs=c(.025,.97))[2])),"yes","no"))
a2 <- filter(a, GHTime == 0)
a2["Unhealthy"] <- "yes"
a <- rbind(a,a2)

a$Unhealthy <- as.factor(a$Unhealthy)

a0 <- filter(a, Unhealthy == "yes")
a0 %>% arrange(Label) 

pred.elev.synd <- GAMBuild(a0,"Reduced Range")

a1 <- filter(a, Unhealthy == "no")
a1 %>% arrange(Label) 

pred.norm.synd <- GAMBuild(a1,"Normal Range")

pred.synd <-rbind(pred.elev.synd,pred.norm.synd)


synd.max.plus2<-quantile(subset(a,GHTime==0,var)[,1],c(0.975))
synd.min.plus2<-quantile(subset(a,GHTime==0,var)[,1],c(0.025))

PlotGH <- PlotGH %>% mutate(Third = 
  case_when(Label == "Healthy" ~ "Healthy controls", 
            Unhealthy == "no" ~ "Non-reduced trauma patients", 
            TRUE ~ "Reduced trauma patients")
)



synd.plot2<-ggplot(data=pred.synd,aes(x=Time,y=Fit,group=Syndecan,fill=Syndecan)) + theme_bw(base_size=8) + ggtitle(paste0(i)) + xlab('Time (minutes)') + ylab(i)
synd.plot2<-synd.plot2 + geom_ribbon(aes(ymax=synd.max.plus2,ymin=synd.min.plus2,fill="Healthy Limits"),col=NA,alpha=0.05) 
synd.plot2<-synd.plot2 + geom_ribbon(aes(ymin=Lower,ymax=Upper),alpha=0.25) + geom_line(size=1) + scale_x_continuous(limits = c(0,maxVal), breaks=seq(0,maxVal,by=10), labels=seq(0,maxVal,by=10),expand = c(0.01, 0))#+ ylim(-1,60) #explore this more
synd.plot2<-synd.plot2 + geom_point(data=PlotGH,aes(x=GHTime,y=var,group=Third,colour=Third),shape=21,size=1,fill='black',stroke=0.8)

#Which days sees the mean predicted Syndecan value in elevated patients cross over the upper 2.5th percentile of the distribution of 'Normal' values?
low.lim<-pred.elev.synd$Time[which(pred.elev.synd$Lower<synd.min.plus2)[[1]]]
up.lim<-pred.elev.synd$Time[which(pred.elev.synd$Upper<synd.min.plus2)[[1]]]
est.ave<-pred.elev.synd$Time[which(pred.elev.synd$Fit<synd.min.plus2)[[1]]]
StartingTime <- c(est.ave,low.lim,up.lim)
est.line<-data.frame(x.val=c(est.ave,est.ave),y.val=c(0,synd.min.plus2))


synd.plot2 <- synd.plot2 + 
  geom_vline(aes(xintercept=est.line$x.val[1],linetype='Estimated time of onset'), col = "black",lwd=0.7) + scale_color_manual("", values = c("dark green", "blue", "dark orange"), labels = c("Healthy controls", "Trauma: Non-reduced levels", "Trauma: Reduced levels")) + scale_linetype_manual("", values = "dashed") + scale_fill_manual("", values = c("green", "blue", "red"), labels = c( "Healthy baseline range",  "Trauma: Non-reduced range", "Trauma: Reduced range")) + guides(fill= guide_legend(title = ""))  + theme(legend.text=element_text(size=10))


    return(list(synd.plot2,StartingTime))
}

```



```{r warning=FALSE, include=FALSE}

ChosenData <- AllDataLOQLong %>% 
  filter(variable != "Aldosterone") #added September

ErrorPlots <- list()


pQuant <- list()
pQuant2 <- list()
StartingTime <- list()
pQuanta <- list()
pQuant2a <- list()
StartingTimea <- list()

variableInt <- ChosenData %>% 
  filter(Descriptor == "Steroid") %>% 
  select(variable) %>% 
  unique() 

for (i in  variableInt$variable){ 
    
    tryCatch({
    
      print(i)
    
PlotGH <- filter(ChosenData, variable %in% c("GoldenHour", i), Time %in% c("Clinical", "T1","HC")) %>% 
  select(ID, Label, variable, value,Time) 

PlotGH <- dcast(PlotGH, Label +ID   ~ variable ,na.rm = TRUE)

PlotGH[which(PlotGH$Label == "Healthy Control"), "GoldenHour" ] <- 0
PlotGH <- PlotGH[,-c(2)]
PlotGH <-    PlotGH %>% 
  mutate_if(names(.) %in% c(i, "GoldenHour"),as.numeric) #%>% 
  #filter(GoldenHour < 51)
PlotGH$Label <- as.factor(PlotGH$Label)

levels(PlotGH$Label) <- c("Healthy", "Trauma", "Trauma")

        if (i %in% c("Progesterone","Sex","Testosterone","Testosterone/Cortisol")){
        
            print("Yes")
             names(PlotGH) <- c("Label","GHTime","var")

            }else{
            
            names(PlotGH) <- c("Label","var","GHTime")
}


if ((mean(filter(PlotGH, Label == "Healthy")$var)) <= (mean(filter(PlotGH, Label == "Trauma")$var ,na.rm=TRUE))){

  print(paste0(i,"-elevated"))
  PlotGH <- mutate(PlotGH, Unhealthy = ifelse(var >= as.numeric((quantile(filter(PlotGH, Label == "Healthy")$var,probs=c(.025,.97))[2])),"yes","no"))   
 
  pQuant2a[[i]] <- ggplot(PlotGH, aes( GHTime,var, colour=Unhealthy))+geom_point()+labs(y= paste(i))+ggtitle(paste0(i))+theme_bw()

  resulta <- GAMModel(PlotGH,i)
}
  else{
   
    print(paste0(i,"-reduced"))
    PlotGH <- mutate(PlotGH, Unhealthy = ifelse(var <= as.numeric((quantile(filter(PlotGH, Label == "Healthy")$var,probs=c(.025,.97))[1])),"yes","no"))   
   
   pQuant2a[[i]] <- ggplot(PlotGH, aes( GHTime,var, colour=Unhealthy))+geom_point()+labs(y= paste(i))+ggtitle(paste0(i)) +theme_bw()

    resulta <- GAMModel2(PlotGH,i)
    
  }
    
  pQuanta[[i]] <- resulta[[1]] + theme(legend.title=element_blank())
    StartingTimea[[i]] <- resulta[[2]]
 
}, error=function(e){
  
  
  cat("ERROR :",conditionMessage(e), "\n")})
}


```

### Golden Hour GAMs

```{r include=FALSE}

pdf("ConorFig8FinalLow.pdf", height = 10, width = 25)

wrap_plots(pQuanta) + plot_layout(guides = "collect") 

dev.off()


pdf("ConorFig8FinalAllLow.pdf", height = 10, width = 25)

wrap_plots(pQuant2a)  + plot_layout(guides = "collect") 

dev.off()
```

```{r GAMs for steroids in first time point (golden hour), fig.cap = "Golden Hour GAM", fig.width= 15, fig.height= 9}

wrap_plots(pQuanta)   + plot_layout(guides = "collect")

```
Original plots with patients above/below normal healthy concentrations. GAMs tries to build two different models (around blue dots and orange dots). As you can see, those steroids without model are those where the separation between orange and blue is practically non existent (e.g DHEA/DHEAS)

```{r Steroids in first time point (golden hour), fig.cap = "Golden Hour All", fig.width= 12, fig.height= 9}

wrap_plots(pQuant2a)  + plot_layout(guides = "collect") 

```

## Heatmaps

```{r}
Mineralocorticoids <- c("11-deoxycorticosterone","Corticosterone","Aldosterone")
GlucocorticoidPrec <- c("Progesterone","17-hydroxyprogesterone","11-deoxycortisol")
Glucocorticoids <- c("Cortisol", "Cortisone" )
Androgens <- c("DHEA","DHEAS","Androstenedione")
OxygenatedAndro <- c("Testosterone", "Dihydrotestosterone","11-KA4","11-KTestosterone","11B-OHA4","11B-OHT")
Ratios <- c("Cortisol/Cortisone", "Cortisol/DHEA","Cortisol/DHEAS","DHEA/DHEAS","DHEA/Testosterone","Testosterone/Cortisol")

aa <- list(Mineralocorticoids,GlucocorticoidPrec, Glucocorticoids,Androgens, OxygenatedAndro, Ratios) 
names(aa) <- c("Mineralocorticoids","GlucocorticoidPrec", "Glucocorticoids","Androgens", "OxygenatedAndro", "Ratios")

ll <- data.frame(aa) %>% 
  pivot_longer(everything()) %>% 
  unique() %>%
  as.data.frame()
```


<https://github.com/kassambara/ggpubr/issues/65> p value adjust

```{r }
# it was working before and now not - issues with names. 

scale_this <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}


Heatmapping <- function(ChosenData,ll){

AverageHC <-  ChosenData %>% 
  filter(Time =="HC") %>%
  group_by(variable) %>%
  mutate(MeanValue = mean(as.numeric(value))) %>% 
  select(variable, MeanValue) %>%
  unique() 

#AverageHC$variable <- str_replace(AverageHC$variable,  "11β-OHA4", "11B-OHA4")
#AverageHC$variable <- str_replace(AverageHC$variable,  "11β-OHT", "11B-OHT")

MixingHM <- inner_join(ChosenData, AverageHC) #adds a column with mean HC value

FinalDataHM <- MixingHM %>% 
  filter( Descriptor %in% c("Steroid"), Time !="HC") 

FinalDataHM$variable <- str_replace(FinalDataHM$variable,  "11β-OHA4", "11B-OHA4")
FinalDataHM$variable <- str_replace(FinalDataHM$variable,  "11β-OHT", "11B-OHT")

FinalDataHM$variable <- droplevels(as.factor(FinalDataHM$variable))
FinalDataHM$value <- as.numeric(FinalDataHM$value)

FinalDataHM <- FinalDataHM %>%
  mutate(valueWHC = (value-MeanValue)) %>%
  group_by(variable) %>%
  mutate(value_scaled = scale_this(valueWHC)) %>%
  select(ID, variable, value_scaled, Time) %>%
  pivot_wider(id_cols = ID, names_from = c(variable, Time), values_from = value_scaled) 

Clinical <- ChosenData %>% 
  filter( Descriptor !=  c("Steroid"), Time !="HC") %>% 
  select(ID, variable, value, Time) %>%
  pivot_wider(id_cols = ID, names_from = c(variable, Time), values_from = value) %>% 
  as.data.frame()

FinalDataHMID <- FinalDataHM$ID
FinalDataHMMatrix <- as.matrix(FinalDataHM %>% 
                      select(-c(ID, '11B-OHT_T3','11B-OHT_T2'))) ## too many missing values
rownames(FinalDataHMMatrix) = FinalDataHMID

ll3 <- data.frame(str_split(colnames(FinalDataHMMatrix), "_", simplify = TRUE))
names(ll3) <- c("value", "time")
ll2 <- inner_join(ll3,  ll) 


annotation_col = data.frame(
   
    Time = ll2$time,
    Type  = ll2$name 
    #Mean = ll2a$Mean
)
rownames(annotation_col) = colnames(FinalDataHMMatrix)


annotation_row = data.frame(
    
    Age= as.numeric(Clinical$Age_Clinical),
    LOS = as.numeric(Clinical$LOS_Clinical),
    ISS = as.numeric(Clinical$ISS_Clinical),
    Fluids = as.factor(Clinical$Fluids_Clinical),
    Anaesthesia = as.factor(Clinical$Anaesthetic_Clinical)
)
rownames(annotation_row) = Clinical$ID


ann_colors = list(
  
  Type = c(Androgens = "light blue", GlucocorticoidPrec = "yellow", Glucocorticoids = "orange", Mineralocorticoids = "springgreen4",OxygenatedAndro= "dark blue", Ratios ="snow2"),
  Time = c(T1 = "green", T2 = "orange", T3 = "red"), 
  Anaesthesia = c(NO = "lawngreen", YES = "red2"), 
  Fluids = c(NO = "lawngreen", YES = "red2")#, 
  #Age =  c("white", "firebrick"),
  #ISS = c("white", "firebrick"),
  #LOS = c("white", "firebrick")
)


#cutree_cols  = 5,cutree_rows  = 3,
Heatmap <- pheatmap(FinalDataHMMatrix,  annotation_col = annotation_col, annotation_row = annotation_row, cutree_cols  = 5,cutree_rows  = 3,annotation_colors = ann_colors, fontsize_col = 9, labels_row =paste0("Patient.", 1:31))#, fontsize_row = 0


   # annotation_colors = ann_colors)
return(list(FinalDataHM, ll,Clinical, Heatmap ))
}
```

Once again, main difference with original LOQ data is the amount of missing values (grey cells) [ Had to delete '11B-OHT_T3','11B-OHT_T2' as had only one value left]


```{r Heatmap all time points, fig.cap = "Heatmap with NO LOQ change", fig.height= 9, fig.width= 14}

AldNo <- AllDataLOQLong %>% 
  filter(variable != "Aldosterone")
Heatmapping(AldNo,ll)[[4]]

pdf("HeatmapAll.pdf", 17, 11)
Heatmapping(AldNo,ll)[[4]] #too many missing values!
dev.off()

```

### General average

```{r Average steroid concentration through time, fig.cap = "Heatmap average", fig.width= 8, fig.height= 6}


# it was working before and now not - issues with names. 

scale_this <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}



Average <-  AldNo %>% 
  filter(Time != "Clinical") %>%
  group_by(variable) %>%
  mutate(value_scaled = scale_this(as.numeric(value))) %>%
  ungroup() %>%
  group_by(variable,Time) %>%
  mutate(MeanValue = mean(as.numeric(value_scaled))) %>% 
  select(variable, MeanValue) %>%
  unique() 

Average$variable <- str_replace(Average$variable,  "11β-OHA4", "11B-OHA4")
Average$variable <- str_replace(Average$variable,  "11β-OHT", "11B-OHT")

AverageMat <- pivot_wider(Average,names_from = Time, values_from =MeanValue ) 

FinalDataHMID <- AverageMat$variable
AverageMatFinal <- AverageMat %>% as.data.frame() %>%
  dplyr::select(c("HC","T1", "T2", "T3"))
FinalDataHMMatrix <- as.matrix(AverageMatFinal)
rownames(FinalDataHMMatrix) <- FinalDataHMID

ll2 <- ll[match(rownames(FinalDataHMMatrix), ll$value),] %>% 
  filter(value != "Aldosterone") 

annotation_row = data.frame(
    
    Type  = ll2$name
)


rownames(annotation_row) = rownames(FinalDataHMMatrix) 


ann_colors = list(
  
  Type = c(Androgens = "light blue", GlucocorticoidPrec = "yellow", Glucocorticoids = "orange", Mineralocorticoids = "springgreen4",OxygenatedAndro= "dark blue", Ratios ="snow2")
)

######
AveragePlot <- pheatmap(FinalDataHMMatrix,  annotation_row = annotation_row, annotation_colors = ann_colors, fontsize_col = 9,cluster_cols=FALSE)#,

AveragePlot

pdf("HeatmapAverageConce.pdf",  7,  5)
print(AveragePlot)
dev.off()

```





